---
title: "Automating antibiotic treatment choice using clinical prediction modelling: a microsimulation study"
format:
  docx:
    reference-doc: doc_template.docx
bibliography: UF_references.bib
csl: /Users/alexhoward/Zotero/styles/the-lancet-infectious-diseases.csl
editor: visual
---


```{r, echo=FALSE,message=FALSE,warning=FALSE}

library(tidyverse)
library(knitr)
library(glue)

```

# Supplementary Material

## Appendix 1: Discrete choice experiment questionnaire

```{r}

print("Screenshot of discrete choice experiment here")

```
Figure 1: The discrete choice experiment provided to prescribers

## Appendix 1: Data processing workflow

```{r, echo=FALSE,message=FALSE,warning=FALSE}
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Data flow chart.pdf")
```
Appendix 1: Data processing workflow

## Appendix 2: Utility function

Antibiotic treatment choice utility ($U$) is defined by:

$$
U^{(x)(n)} = w_s^{(x)}v_s^{(x)(n)}\left(  \sum_{k \in K} u\left(w_k, v_k^{(x)(n)}\right) + 
\sum_{l \in L} u\left(w_l^{(x)}, v_l^{(x)(n)}\right) +
u\left(w^{(n)}_i, v_i^{(x)}\right) \right)
$$ {#eq-utility}

where

$$
u(w, v) = 
\begin{cases}
    wp & \text{if } w \geq 0 \\
    |w|(1-v) & \text{otherwise} \\
\end{cases}
$$

$x$ is the antibiotic, $n$ is the patient, $K = \{u, a, o, r, h\}$, $L = \{c, t\}$, $v$ are values, and $w$ are weights. Weights $w_k$ in $K$ are defined solely by corresponding ranked logit coefficients from the clinician discrete choice experiment (DCE). Weight $w_s$ is defined solely by the area under the receiver operating characteristic (AUROC) value for the susceptibility clinical prediction model on the model validation dataset. Weights $w_l$ in $L$ are defined by:

$$
w_l^{(x)} = w_{l,1} + w_{l,2}^{(x)}
$$

where $w_{l,1}$ is the corresponding DCE ranked logit coefficient and $w_{l,2}$ is the validation AUROC value for the respective clinical prediction model.

$w_i$ is defined by:

$$
w_i^{(n)} = w_{i,1} + w_{i,2}^{(n)}
$$

where $w_{i,1}$ is the corresponding DCE ranked logit coefficient and $w_{i,2}$ is an illness severity weight defined by the user.

Values are defined as follows:

$$
v_s \in [0,1] \quad \text{(Probability of organism antimicrobial susceptibility)}
$$

$$
v_u \in \{0,1\} \quad \text{(UTI-specific agent)}
$$

$$
v_a \in \{0,1\} \quad \text{(WHO Access agent)}
$$

$$
v_o \in \{0,1\} \quad \text{(Orally-administrable agent)}
$$

$$
v_i \in \{0,1\} \quad \text{(Intravenously-administrable agent)}
$$

$$
v_r \in \{0,1\} \quad \text{(WHO Reserve agent)}
$$

$$
v_h \in [0,1] \quad \text{(Normalised agent cost)}
$$

$$
v_c \in [0,1] \quad \text{(Probability of } \textit{Clostridiodes difficile} \text{ infection)}
$$

$$
v_t \in [0,1] \quad \text{(Probability of agent toxicity)}
$$

## Appendix 2

```{r, echo=FALSE,message=FALSE,warning=FALSE}

desc_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/uf_desctab.csv")

desc_tab$Characteristic[is.na(desc_tab$Characteristic)] <- "··"

knitr::kable(desc_tab)

```
Appendix 2: Patient compositions of the model development and microsimulation datasets.

##Appendix 3

```{r, echo=FALSE,message=FALSE,warning=FALSE}

ab_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/ab_tab.csv")

knitr::kable(ab_tab,)

```
Appendix 3:: Antibiotic courses in the prescription dataset, including two-combination administrations of more than 24 hours duration.

## Appendix 4

```{r, echo=FALSE,message=FALSE,warning=FALSE}

hyp_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/hyp_tab.csv")

knitr::kable(hyp_tab)

```
Appendix 4: Final training hyperparameters used for the 39 clinical prediction models.

## Appendix 5

```{r, echo=FALSE,message=FALSE,warning=FALSE}

feat_table <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/feat_table.csv")

knitr::kable(feat_table)

```
Appendix 5: Feature importances in the clinical prediction models.

##Appendix 6

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/roc_plots_other.pdf")
  
```
Appendix 6: ROC curves for probability predictions made for 2-antibiotic combination susceptibility on the validation dataset.

## Appendix 7

```{r, echo=FALSE,message=FALSE,warning=FALSE}

matrics_combos_table <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/metrics_combos_table.csv")

knitr::kable(matrics_combos_table)

```
Appendix 7: Model performance metrics for 2-antibiotic combinations in the single validation run on the validation dataset.


## Appendix 8

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Accuracy.pdf")
  
```
Appendix 8: Small dataset stability metrics for single antibiotic, CDI, and toxicity models in six random train-test splits stratified by outcome at a sequence of small dataset sizes across 50 training runs of XGBoost.

## Appendix 9

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Accuracy.pdf")

```
Appendix 9: Year group cluster analysis of clinical prediction model performance in six random train-test splits stratified by outcome.

## Appendix 10

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Accuracy.pdf")

  
```
Appendix 10: Fairness analyses of clinical prediction model performance across age, race, gender, spoken language, and marital status in six random train-test splits stratified by outcome.

## Appendix 11

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Medicine_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Surgery_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Infection_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Intensive care_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/General Practice_importances.pdf")

```
Appendix 11: Coefficients from the ranked logit analysis of the discrete choice experiment across all prescriber specialties

## Appendix 12

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, medical specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, surgical specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, infection specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, intensive care).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, general practice).pdf")

```
Appendix 12: The distribution of utility values across the microsimulation dataset for all agents (including 2-inpatient combinations), both overall and stratified by the individual specialities

## Appendix 13

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_.pdf")

```
Appendix 13: The overall distribution of utility values across the microsimulation dataset for all antibiotics including 2-agent combinations.

## Appendix 14

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/UTI-specificity_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Access category_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Oral option_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/IV option_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Cost_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Reserve category_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/CDI_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Toxicity_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/auc_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/resistance_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/CDI probability_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/toxicity probability_sens_plots_grid.pdf")


```
Appendix 14: Sensitivity analyses demonstrating the effect of varying all utility function parameters on overall treatment utility of individual antibiotics.
