---
title: "Automating antibiotic treatment choice using clinical prediction modelling: a microsimulation study"
format:
  docx:
    reference-doc: doc_template.docx
bibliography: UF_references.bib
csl: /Users/alexhoward/Zotero/styles/the-lancet-infectious-diseases.csl
editor: visual
---

```{r, echo=FALSE,message=FALSE,warning=FALSE}

library(tidyverse)
library(knitr)
library(glue)

```

# Appendix

## Appendix 1: Discrete choice experiment questionnaire

This survey is designed to ascertain the relative importance of different features of an antimicrobial to clinicians when prescribing. It should take 2-5 minutes to complete. Contextual information about the patient has intentionally been omitted - your decision should be made based on your experience of managing urinary tract infection in your usual clinical setting. The factors listed are AWaRe classification\*, risk of Clostridioides difficile diarrhoea, toxicity risk (e.g., nephrotoxicity), UTI-specificity (whether the drug should only be used for UTI), oral option, intravenous option, and financial cost.

-   The WHO AWaRe classification assigns antibiotics ratings of Access, Watch and Reserve based on their risk of generating antimicrobial resistance. Those in the Access category are the lowest risk, those in Reserve are the highest.

A patient has urinary tract infection. You have no other details about the patient at this stage (including disease severity). Please rank the following fictional antibiotic treatments from 1 to 13 in order of most to least preferred in this scenario by dragging and dropping options or using the arrows on the right. When finished, just click the 'Done' button at the bottom of the page.

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics(c(
  "/Users/alexhoward/Documents/Projects/UDAST_code/S1.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S2.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S3.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S4.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S5.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S6.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S7.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S8.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S9.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S10.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S11.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S12.png",
  "/Users/alexhoward/Documents/Projects/UDAST_code/S13.png"
                          ))
                      

```

Appendix 1: A replica of the discrete choice experiment provided to prescribers\

## Appendix 2: Data processing workflow

```{r, echo=FALSE,message=FALSE,warning=FALSE}
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Data flow chart.pdf")
```

Appendix 2: Data processing workflow\

## Appendix 3: Model specification and training details

XGBoost models were trained and tuned on the training datasets using L2 regularisation (ridge penalty) to control overfitting. Area under the receiver operating characteristic curve (AUROC) was used as the model evaluation metric. Hyperparameter tuning was performed sequentially for each of the 39 models across parameters in 3 stages. 10 sets of paired values for maximum tree depth (selecting values between two and nine inclusive) and minimum child weight (selecting values between one and 10 inclusive) were randomly selected using Latin hypercube sampling with 'randomLHS' from the 'lhs' package (<https://cran.r-project.org/web/packages/lhs/index.html>) - a 5-fold cross-validation was then performed across 50 boosting rounds for each of these 10 pairs of values, with a default learning rate of 0.05, a proportion of cases used for each decision tree (subsample rows ratio) of 0.8, and a proportion of predictors used for each decision tree (subsample columns ratio) of 0.8.[@carnellLhsLatinHypercube2006]

The maximum tree depth and minimum child weight values that produced the highest AUROC were carried forward to another set of 5-fold cross-validations with 50 boosting rounds across 10 paired subsample rows and columns ratios (again randomly selected using 'randomLHS', with both values sampling from a range between 0.5 and 1.0 inclusive) with a learning rate of 0.05. The subsample rows and columns ratios that produced the highest AUROC were carried forward to a final set of 5-fold cross-validations across four learning rates (0.1, 0.05, 0.01, and 0.001), with a maximum number of boosting rounds of 1000 and early stopping if AUROC did not improve for 50 rounds.

The learning rate and the round number at which the best AUROC value occurred (forming the new total number of rounds parameter) were then taken forward as parameters for training the final model. Class imbalance methods (e.g., class weighting) were not used to ensure that probability predictions were not skewed to improve predictive performance (e.g., universally increased to improve sensitivity), given that they would subsequently be used to inform the utility function.

Predictor variable contributions to predictive value in model training were measured using Shapley values with 'predict' from the 'stats' package.[@Chapter53Shapley] Predictor variables with Shapley values of zero were then excluded from models for the final training round.

## Appendix 4: Utility function

Antibiotic treatment choice utility ($U$) is defined by:

$$
U^{(x)(n)} = w_s^{(x)}v_s^{(x)(n)}\left(  \sum_{k \in K} u\left(w_k, v_k^{(x)(n)}\right) + 
\sum_{l \in L} u\left(w_l^{(x)}, v_l^{(x)(n)}\right) +
u\left(w^{(n)}_i, v_i^{(x)}\right) \right)
$$ {#eq-utility}

where

$$
u(w, v) = 
\begin{cases}
    wp & \text{if } w \geq 0 \\
    |w|(1-v) & \text{otherwise} \\
\end{cases}
$$

$x$ is the antibiotic, $n$ is the patient, $K = \{u, a, o, r, h\}$, $L = \{c, t\}$, $v$ are values, and $w$ are weights. Weights $w_k$ in $K$ are defined solely by corresponding ranked logit coefficients from the clinician discrete choice experiment (DCE). Weight $w_s$ is defined solely by the area under the receiver operating characteristic (AUROC) value for the susceptibility clinical prediction model on the model validation dataset. Weights $w_l$ in $L$ are defined by:

$$
w_l^{(x)} = w_{l,1} + w_{l,2}^{(x)}
$$

where $w_{l,1}$ is the corresponding DCE ranked logit coefficient and $w_{l,2}$ is the validation AUROC value for the respective clinical prediction model.

$w_i$ is defined by:

$$
w_i^{(n)} = w_{i,1} + w_{i,2}^{(n)}
$$

where $w_{i,1}$ is the corresponding DCE ranked logit coefficient and $w_{i,2}$ is an illness severity weight defined by the user.

Values are defined as follows:

$$
v_s \in [0,1] \quad \text{(Probability of organism antimicrobial susceptibility)}
$$

$$
v_u \in \{0,1\} \quad \text{(UTI-specific agent)}
$$

$$
v_a \in \{0,1\} \quad \text{(WHO Access agent)}
$$

$$
v_o \in \{0,1\} \quad \text{(Orally-administrable agent)}
$$

$$
v_i \in \{0,1\} \quad \text{(Intravenously-administrable agent)}
$$

$$
v_r \in \{0,1\} \quad \text{(WHO Reserve agent)}
$$

$$
v_h \in [0,1] \quad \text{(Normalised agent cost)}
$$

$$
v_c \in [0,1] \quad \text{(Probability of } \textit{Clostridiodes difficile} \text{ infection)}
$$

$$
v_t \in [0,1] \quad \text{(Probability of agent toxicity)}
$$

## Appendix 5

```{r, echo=FALSE,message=FALSE,warning=FALSE}

desc_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/uf_desctab.csv")

desc_tab$Characteristic[is.na(desc_tab$Characteristic)] <- "··"

knitr::kable(desc_tab)

```

Appendix 5: Patient compositions of the model development and microsimulation datasets.\

## Appendix 6: Antibiotic courses in the prescription dataset

```{r, echo=FALSE,message=FALSE,warning=FALSE}

ab_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/ab_tab.csv")

knitr::kable(ab_tab,)

```

Appendix 6: Antibiotic courses in the prescription dataset, including two-combination administrations of more than 24 hours duration.\

## Appendix 7: Final training hyperparameters

```{r, echo=FALSE,message=FALSE,warning=FALSE}

hyp_tab <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/hyp_tab.csv")

knitr::kable(hyp_tab)

```

Appendix 7: Final training hyperparameters used for the 39 clinical prediction models.\

## Appendix 8: Feature importances

```{r, echo=FALSE,message=FALSE,warning=FALSE}

feat_table <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/feat_table.csv")

knitr::kable(feat_table)

```

Appendix 8: Feature importances in the clinical prediction models.\

## Appendix 9: ROC curves for 2-antibiotic combination models

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/roc_plots_other.pdf")
  
```

Appendix 9: ROC curves for probability predictions made for 2-antibiotic combination susceptibility on the validation dataset.\

## Appendix 10: Model performance metrics for 2-antibiotic combinations

```{r, echo=FALSE,message=FALSE,warning=FALSE}

metrics_combos_table <- read_csv("/Users/alexhoward/Documents/Projects/UDAST_code/metrics_combos_table.csv")

knitr::kable(metrics_combos_table)

```

Appendix 10: Model performance metrics for 2-antibiotic combinations in the single validation run on the validation dataset.\

## Appendix 11: Small dataset stability metrics

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/stability_Accuracy.pdf")
  
```

Appendix 11: Small dataset stability metrics for single antibiotic, CDI, and toxicity models in six random train-test splits stratified by outcome at a sequence of small dataset sizes across 50 training runs of XGBoost.\

## Appendix 12: Year group cluster analysis

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2008 - 2010_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2011 - 2013_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2014 - 2016_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/timesens_2017 - 2019_Accuracy.pdf")

```

Appendix 12: Year group cluster analysis of clinical prediction model performance in six random train-test splits stratified by outcome.\

## Appendix 13: Fairness analyses

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Age group_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_Race_Accuracy.pdf")

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_AUC-ROC.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Precision.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Recall.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_F1 score.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/protchar_(Gender|Language|Marital)_Accuracy.pdf")

  
```

Appendix 13: Fairness analyses of clinical prediction model performance across age, race, gender, spoken language, and marital status in six random train-test splits stratified by outcome.\

## Appendix 13: Coefficients from the ranked logit analysis

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Medicine_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Surgery_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Infection_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Intensive care_importances.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/General Practice_importances.pdf")

```

Appendix 14: Coefficients from the ranked logit analysis of the discrete choice experiment across all prescriber specialties.\

## Appendix 15: The distribution of utility values by specialty

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, medical specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, surgical specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, infection specialties).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, intensive care).pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_ (single agent, general practice).pdf")

```

Appendix 15: The distribution of utility values across the microsimulation dataset for all single agents, stratified by the individual specialities.\

## Appendix 16: The distribution of utility values for including antibiotic combinations

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/utility_Treatment_.pdf")

```

Appendix 16: The overall distribution of utility values across the microsimulation dataset for all antibiotics including 2-agent combinations.\

## Appendix 17: Sensitivity analysis of utility calculation

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/UTI-specificity_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Access category_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Oral option_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/IV option_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Cost_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Reserve category_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/CDI_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Toxicity_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/auc_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/CDI_auc_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/Toxicity_auc_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/resistance_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/CDI probability_sens_plots_grid.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/toxicity probability_sens_plots_grid.pdf")


```

Appendix 17: Sensitivity analyses demonstrating the effect of varying all utility function parameters on overall treatment utility of individual antibiotics.\

## Appendix 18: Number of susceptible results in the top six recommendations

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/UF_All agents_WHO access agents_plot.pdf")

```

Appendix 18: The number of susceptible results in the top six automated recommendations (PDRx), overall and for WHO Access agents, compared to a standard panel of nitrofurantoin, trimethoprim-sulfamethoxazole, gentamicin, piperacillin-tazobactam, ceftriaxone, and ciprofloxacin. Median values are displayed as red/green dots, interquartile ranges as red/green lines, and all results as grey dot cloud darkness.\

## Appendix 19: Illness severity sensitivity analysis (second-line agents)

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of WHO Access agents, 2nd-line)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of orally-administrable agents, 2nd-line)_susplot.pdf")
knitr::include_graphics(normalizePath("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of IV-administrable agents, 2nd-line)_susplot.pdf"))
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/ilness_2ndplot.pdf")

```

Appendix 19: Changes in the proportion of urine organism coverage with second-line treatment recommendations overall, Access treatment recommendations, treatment recommendations with oral preparations, and treatment recommendations with intravenous preparations as the illness severity of the population was increased.\

## Appendix 20: Illness severity sensitivity analysis (combinations included)

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of WHO Access agents, combinations included)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of orally-administrable agents, combinations included)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of IV-administrable agents, combinations included)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/ilness_comboplot.pdf")

```

Appendix 20: Changes in the proportion of urine organism coverage with first-line treatment recommendations overall, Access treatment recommendations, treatment recommendations with oral preparations, and treatment recommendations with intravenous preparations as the illness severity of the population was increased (2-agent combinations included).\

## Appendix 21: Illness severity sensitivity analysis (improved probability predictions)

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of WHO Access agents, improved probability predictions)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of orally-administrable agents, improved probability predictions)_susplot.pdf")
knitr::include_graphics(normalizePath("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to illness severity (proportion of IV-administrable agents, improved probability predictions)_susplot.pdf"))
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/ilness_impplot.pdf")

```

Appendix 21: Changes in the proportion of urine organism coverage with first-line treatment recommendations overall, Access treatment recommendations, treatment recommendations with oral preparations, and treatment recommendations with intravenous preparations as the illness severity of the population was increased (improved probability predictions).\

## Appendix 22: Illness severity sensitivity analysis (nitrofurantoin resistance)

```{r, echo=FALSE,message=FALSE,warning=FALSE}

knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to median nitrofurantoin resistance probability (proportion of WHO Access agents)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to median nitrofurantoin resistance probability (proportion of orally-administrable agents)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/overall _MEWS_according to median nitrofurantoin resistance probability (proportion of IV-administrable agents)_susplot.pdf")
knitr::include_graphics("/Users/alexhoward/Documents/Projects/UDAST_code/nitres_abplot.pdf")

```

Appendix 22: Changes in the proportion of urine organism coverage with first-line treatment recommendations overall, Access treatment recommendations, treatment recommendations with oral preparations, and treatment recommendations with intravenous preparations as the median predicted probability of nitrofurantoin resistance was varied between zero and one.\

## References

::: {#refs}
:::


